Trilemma

### 2022-02-11 Trilemma_clustering.Rmd
improved figures for paper

---
title: "Trilemma"
author: "Juan Rocha"
date: "Updated `r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
editor_options: 
  chunk_output_type: inline
---

@@

# # clustering
# library(vegan)
# library(NbClust)
# library(clValid)
# library(mclust)

@@

```{r}
p1 <- df_dat %>% 
    ggplot(aes(EFconsPerCap, gini, group = country)) +
    geom_point(size = 0.5) +
    geom_path(size = 0.2, aes(color = year)) +
    annotate(geom = "rect", xmin = 0, xmax = 1.68, ymin = 20, ymax = 30,
             fill = "skyblue", alpha = 0.25) +
    annotate(geom ="rect",xmin = 0, xmax = 1.68, ymin = 30, ymax = 62,
             fill = "orange", alpha = 0.1) +
    annotate(geom ="rect", xmin = 1.68, xmax = 18, ymin = 20, ymax = 30,
             fill = "orange", alpha = 0.1) +
    annotate(geom="rect", xmin = 1.68, xmax = 18, ymin = 30, ymax = 62,
             fill = "red", alpha = 0.1) +
    geom_hline(yintercept = 30, linetype = 2, color = "gray50") +
    geom_vline(xintercept = 1.68, linetype = 2, color = "gray50") +
    scico::scale_color_scico("Year", palette = "bilbao") +
    labs(y = "Gini", x = "Ecological footprint") + theme_light() +
    theme(legend.position = "top", panel.grid = element_blank(),
          legend.key.width = unit(10, "mm"), legend.key.height = unit(3,"mm"))

p2<- df_dat %>% 
  ggplot(aes(gni2, gini, group = country)) +
  geom_point(size = 0.5, show.legend = FALSE) +
  geom_path(size = 0.2, aes(color = year)) +
    annotate(geom = "rect", xmin = 10, xmax = 1.68, ymin = 20, ymax = 30,
             fill = "skyblue", alpha = 0.25) +
    annotate(geom ="rect",xmin = 0, xmax = 1.68, ymin = 30, ymax = 62,
             fill = "orange", alpha = 0.1) +
    annotate(geom ="rect", xmin = 1.68, xmax = 18, ymin = 20, ymax = 30,
             fill = "orange", alpha = 0.1) +
    annotate(geom="rect", xmin = 1.68, xmax = 18, ymin = 30, ymax = 62,
             fill = "red", alpha = 0.1) +
    geom_hline(yintercept = 30, linetype = 2, color = "gray50") +
    geom_vline(xintercept = 12746, linetype = 2, color = "gray50") +
    scico::scale_color_scico("Year", palette = "bilbao") +
      scale_x_log10() +
    labs(y = "Gini", x = "Ecological footprint") + theme_light() +
    theme(legend.position = "top", panel.grid = element_blank(),
          legend.key.width = unit(10, "mm"), legend.key.height = unit(3,"mm"))

p3<- df_dat %>% 
  ggplot(aes(gni2, EFconsPerCap, group = country)) +
  geom_point(size = 0.5, show.legend = FALSE) +
  geom_path(size = 0.2, aes(color = year)) +
  scale_x_log10()

p1 + p2 + p3 + plot_layout(guides = "collect") & theme_light(base_size = 18)
```

@@

# ggsave(
#     plot = p, path = "figures/", file = "first_diff.png", device = "png",
#     width = 6, height = 3, dpi = 300, bg = "white")
p + theme_light(base_size = 20)
```
```{r fig.width=3, fig.height=3}
df_dat |> 
    ggplot(aes(EFconsPerCap, gini)) +
    geom_hline(yintercept = c(30,40), linetype = 2, color = "gray50") +
    geom_vline(xintercept = 1.68, linetype = 2, color = "gray50") +
    geom_density2d() +
    geom_point(aes(color = gni_log)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico("GNI [log]") +
    labs(y = "Gini", x = "Ecological footprint") +
    theme_light(base_size = 18) +
    theme(legend.position = "top", panel.grid = element_blank())
```
```{r}
df_dat |> 
    ggplot(aes(gni_log, gini)) +
    geom_density2d() +
    geom_point(aes(color = EFconsPerCap)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico(palette = "berlin") +
    theme_light(base_size = 18)
```
```{r}
df_dat |> 
    ggplot(aes(gni_log, EFconsPerCap)) +
    geom_density2d() +
    geom_point(aes(color = gini)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico() +
    theme_light(base_size = 18)
```
Who is the weird trajectory out there, world? No it's Luxemburg

@@

```{r}
df_summary <- df_dat |> 
    as_tibble() |> 
    ungroup() |> 
    group_by(country) |> 
    summarize(m = sum(magni, na.rm = TRUE), 
              t = sum(theta, na.rm = TRUE),
              g = sum(gamma, na.rm = TRUE),
              sd_mag = sd(magni, na.rm = TRUE),
              sd_t = sd(theta, na.rm = TRUE),
              sd_g = sd(gamma, na.rm = TRUE), 
              ef = mean(EFconsPerCap, na.rm = TRUE),
              inq = mean(gini, na.rm = TRUE),
              gni = mean(gni_log, na.rm = TRUE),
              #sd_ef = sd(ef, na.rm = TRUE),
              sd_inq = sd(gini, na.rm = TRUE),
              sd_gni = sd(gni_log, na.rm = TRUE))

@@

It is a bit contested because the numeric indices favours 2 clusters with 6 of them, but 3 and 4 clusters are favoured by 5 metrics each. The Hubert index (which is graphical) favours 4 indexes as well, tieing up with the 2 cluster option.

```{r}
## Stability & Internal validation
stab <- clValid::clValid(
    obj = df_summary %>% as.data.frame() %>% 
        select(where(is.numeric)) %>%
        as.matrix(),
    nClust = c(2:7),
    clMethods = c("hierarchical", "kmeans",  #"som", "model", "diana", "fanny",
                 "sota", "pam", "clara", "agnes"),
    validation = c('stability', "internal"),
    #metric = "manhattan",
    method = "ward",
    verbose = FALSE
) ## Hierarchical is the optimal method

clValid::optimalScores(stab)
```
The stability analysis also supports the choice of either 2 or 9 clusters. Partition around medioids does perform well with 9 clusters, but others are supporting the 2 clusters.

@@

```{r fig.width=7, fig.height=8}
df_summary |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering) |> 
    arrange(desc(m)) |> 
    mutate(country = as_factor(country)) |> 
    mutate(country = fct_reorder2(country, m,g)) |> 
    pivot_longer(cols = m:sd_gni, names_to = "vars", values_to = "vals") |> 
    ggplot(aes(vals, country)) +
    geom_col(aes(fill = as.factor(clus))) +
    facet_grid(clus~vars, scales = "free") +
    theme_light(base_size = 10)
```

```{r}
df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    summarize(
        ef = sum(d_EF, na.rm = TRUE),
        gini = sum(d_gini, na.rm = TRUE),
        gni = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering) |> 
    ggplot(aes(ef, gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = ef, yend = gini, color = as.factor(clus)), arrow = arrow(length = unit(0.5, "cm"))) +
    geom_text(aes(label = country)) +
    theme_light(base_size = 20)
```

```{r}
df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    summarize(
        ef = sum(d_EF, na.rm = TRUE),
        gini = sum(d_gini, na.rm = TRUE),
        gni = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering) |> 
    ggplot(aes(gni, gini)) +
    geom_segment(aes(x = 0, y = 0, xend = gni, yend = gini, color = as.factor(clus)), arrow = arrow(length = unit(0.5, "cm"))) +
    geom_text(aes(label = country)) +
    theme_light(base_size = 20)
```

>>>>

---
title: "Trilemma"
author: "Juan Rocha"
date: "Updated `r Sys.Date()`"
output:
  html_notebook: 
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
editor_options: 
  chunk_output_type: inline
---

@@

# # clustering
# library(vegan)
# library(NbClust)
# library(clValid)
# library(mclust)
# library(ggbiplot)

@@

```{r}
p1 <- df_dat %>% 
    ggplot(aes(EFconsPerCap, gini, group = country)) +
    geom_point(size = 0.5) +
    geom_path(size = 0.2, aes(color = year)) +
    annotate(geom = "rect", xmin = 0, xmax = 1.68, ymin = 20, ymax = 30,
             fill = "skyblue", alpha = 0.2) +
    annotate(geom ="rect",xmin = 0, xmax = 1.68, ymin = 30, ymax = 62,
             fill = "orange", alpha = 0.2) +
    annotate(geom ="rect", xmin = 1.68, xmax = 18, ymin = 20, ymax = 30,
             fill = "orange", alpha = 0.2) +
    annotate(geom="rect", xmin = 1.68, xmax = 18, ymin = 30, ymax = 62,
             fill = "red", alpha = 0.2) +
    geom_hline(yintercept = 30, linetype = 2, color = "gray50") +
    geom_vline(xintercept = 1.68, linetype = 2, color = "gray50") +
    scico::scale_color_scico("Year", palette = "bilbao") +
    labs(y = "Gini", x = "Ecological footprint", tag = "A") + theme_light() +
    theme(legend.position = "top", panel.grid = element_blank(),
          legend.key.width = unit(10, "mm"), legend.key.height = unit(3,"mm"))

p2<- df_dat %>% 
  ggplot(aes(gni2, gini, group = country)) +
  geom_point(size = 0.5, show.legend = FALSE) +
  geom_path(size = 0.2, aes(color = year)) +
    annotate(geom = "rect", xmin = 10, xmax = 12746, ymin = 20, ymax = 30,
             fill = "orange", alpha = 0.2) +
    annotate(geom ="rect",xmin = 10, xmax = 12746, ymin = 30, ymax = 62,
             fill = "red", alpha = 0.2) +
    annotate(geom ="rect", xmin = 12746, xmax = 105000, ymin = 20, ymax = 30,
             fill = "skyblue", alpha = 0.2) +
    annotate(geom="rect", xmin = 12746, xmax = 105000, ymin = 30, ymax = 62,
             fill = "orange", alpha = 0.2) +
    geom_hline(yintercept = 30, linetype = 2, color = "gray50") +
    geom_vline(xintercept = 12746, linetype = 2, color = "gray50") +
    scico::scale_color_scico("Year", palette = "bilbao") +
      scale_x_log10() +
    labs(y = "Gini", x = "GNI [log]", tag = "B") + theme_light() +
    theme(legend.position = "top", panel.grid = element_blank(),
          legend.key.width = unit(10, "mm"), legend.key.height = unit(3,"mm"))

p3<- df_dat %>% 
  ggplot(aes(gni2, EFconsPerCap, group = country)) +
  geom_point(size = 0.5, show.legend = FALSE) +
  geom_path(size = 0.2, aes(color = year))+
    annotate(geom = "rect", xmin = 10, xmax = 12746, ymin = 0, ymax = 1.68,
             fill = "orange", alpha = 0.2) +
    annotate(geom ="rect",xmin = 10, xmax = 12746, ymin = 1.68, ymax = 18,
             fill = "red", alpha = 0.2) +
    annotate(geom ="rect", xmin = 12746, xmax = 105000, ymin = 0, ymax = 1.68,
             fill = "skyblue", alpha = 0.2) +
    annotate(geom="rect", xmin = 12746, xmax = 105000, ymin = 1.68, ymax = 18,
             fill = "orange", alpha = 0.2) +
    geom_hline(yintercept = 1.68, linetype = 2, color = "gray50") +
    geom_vline(xintercept = 12746, linetype = 2, color = "gray50") +
    scico::scale_color_scico("Year", palette = "bilbao") +
      scale_x_log10() +
    labs(y = "Ecological footprint", x = "GNI [log]", tag = "C") + theme_light() +
    theme(legend.position = "top", panel.grid = element_blank(),
          legend.key.width = unit(10, "mm"), legend.key.height = unit(3,"mm"))

# ggsave(
#     plot = p1 + p2 + p3 + plot_layout(guides = "collect") & 
#         theme_light(base_size = 7) & 
#         theme(legend.position = "bottom", panel.grid = element_blank(),
#           legend.key.width = unit(10, "mm"), legend.key.height = unit(3,"mm")),
#     path = "figures/", file = "cube_time.png", device = "png", dpi = 300,
#     width = 7, height = 3, bg = "white"
# )

p1 + p2 + p3 + 
    plot_layout(guides = "collect") & theme_light(base_size = 18) & 
    theme(legend.position = "bottom", panel.grid = element_blank(),
          legend.key.width = unit(20, "mm"), legend.key.height = unit(3,"mm"))
```

@@

# ggsave(
#     plot = p, path = "figures/", file = "first_diff.png", device = "png",
#     width = 6, height = 3, dpi = 300, bg = "white")
p + theme_light(base_size = 20)
```
```{r fig.width=3, fig.height=3}
c1 <- df_dat |> 
    ggplot(aes(EFconsPerCap, gini)) +
    geom_hline(yintercept = 30, linetype = 2, color = "gray50") +
    geom_vline(xintercept = 1.68, linetype = 2, color = "gray50") +
    geom_density2d() +
    geom_point(aes(color = gni_log)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico("GNI [log]") +
    labs(y = "Gini", x = "Ecological footprint") +
    theme_light(base_size = 7) +
    theme(legend.position = "top", panel.grid = element_blank())

c1 + theme_light(base_size = 18)
```
```{r fig.width=3, fig.height=3}
c2 <- df_dat |> 
    ggplot(aes(gni2, gini)) +
    geom_density2d() +
    geom_point(aes(color = EFconsPerCap)) +
    geom_path(aes(group = country), size = 0.15) +
    geom_hline(yintercept = 30, linetype = 2, color = "gray50") +
    geom_vline(xintercept = (12546), linetype = 2, color = "gray50") +
    scico::scale_color_scico("Ecological footprint",palette = "berlin") +
    labs(x= "GNI [log]", y = "Gini") +
    scale_x_log10() +
    theme_light(base_size = 7) +
    theme(legend.position = "top", panel.grid = element_blank())

c2 + theme_light(base_size = 7)
```
```{r fig.width=3, fig.height=3}
c3 <- df_dat |> 
    ggplot(aes(gni2, EFconsPerCap)) +
    geom_density2d() +
    geom_point(aes(color = gini)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico("Gini") +
    labs(x = "GNI", y = "Ecological footprint") +
    scale_x_log10()+
    geom_hline(yintercept = 1.68, linetype = 2, color = "gray50") +
    geom_vline(xintercept = (12546), linetype = 2, color = "gray50") +
    theme_light(base_size = 7) +
    theme(legend.position = "top", panel.grid = element_blank())

c3 + theme_light(base_size = 18)
```
Who is the weird trajectory out there, world? No it's Luxemburg

@@

```{r fig.width=6, fig.height=2}
df_summary <- df_dat |> 
    as_tibble() |> 
    ungroup() |> 
    group_by(country) |> 
    dplyr::summarize(sd_magnitude = sd(magni, na.rm = TRUE),
              sd_theta = sd(theta, na.rm = TRUE),
              sd_gamma = sd(gamma, na.rm = TRUE), 
              EF = mean(EFconsPerCap, na.rm = TRUE),
              Gini = mean(gini, na.rm = TRUE),
              GNI = mean(gni_log, na.rm = TRUE),
              sd_EF = sd(EFconsPerCap, na.rm = TRUE),
              sd_Gini = sd(gini, na.rm = TRUE),
              sd_GNI = sd(gni_log, na.rm = TRUE),
              magnitude_m = mean(magni, na.rm = TRUE), 
              theta_m = mean(theta, na.rm = TRUE),
              gamma_m = mean(gamma, na.rm = TRUE),
              magnitude_a = sum(magni, na.rm = TRUE), 
              theta_a = sum(theta, na.rm = TRUE),
              gamma_a = sum(gamma, na.rm = TRUE))

@@

The Hubert index (which is graphical) favours 4 groups while D index favours 3 groups. Thus in total 9 metrics proposed 2 groups, 9 metrics proposed 3 groups, and all other options are minorities. Let's continue the analysis with 3 to favour diversity.

```{r}
## Stability & Internal validation
stab <- clValid::clValid(
    obj = df_summary %>% as.data.frame() %>% 
        select(where(is.numeric)) %>%
        as.matrix(),
    nClust = c(2:7),
    clMethods = c("hierarchical", "kmeans",  #"som", "model", "diana", "fanny",
                 "sota", "pam", "clara", "agnes"),
    validation = c('stability', "internal"),
    #metric = "manhattan",
    method = "ward.D2",
    verbose = FALSE
) ## Hierarchical is the optimal method

clValid::optimalScores(stab)
```
The stability analysis also supports the choice of either 2 or 7 clusters. Partition around medioids does perform well with 9 clusters, but others are supporting the 2 clusters.

@@

```{r fig.width=7, fig.height=8}
df_summary |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering) |> 
    mutate(country = as_factor(country)) |> 
    mutate(country = fct_reorder2(country, magnitude, Gini)) |> 
    pivot_longer(cols = sd_magnitude:gamma, names_to = "vars", values_to = "vals") |> 
    ggplot(aes(vals, country)) +
    geom_col(aes(fill = as.factor(clus))) +
    facet_grid(clus~vars, scales = "free") +
    theme_light(base_size = 10)
```

```{r}
df_summary |>
    select(-starts_with("country")) |> 
    GGally::ggpairs()
```

```{r fig.width=3, fig.height=3}
pc <- prcomp(df_summary |> 
                 select(-starts_with("country")),
             center = TRUE, scale. = TRUE)
## Fix short names for plotting
df_summary <- df_summary |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique())

pc1 <- ggbiplot::ggbiplot(pc, groups = as.character(stab@clusterObjs$pam$`2`$clustering),
         ellipse = TRUE, labels = df_summary$country_code, labels.size = 2.5,
         varname.size = 4) +
    scale_color_manual("Groups", values = c("orange", "purple")) +
    theme_light(base_size = 18) +
    theme(legend.position = c(0.2, 0.2), panel.grid = element_blank())

pc1
```

Note to self: It does not make sense to plot the average distance traversed in the space per country. Take Luxemburg as example, it started and ended verly close by. The absolute distance is what I'm calculating with the euclidian distance on that plane, then the first-diff values are just increases or decreases with respect to initial conditions. Averages are then misleading, because on average Luxemburg moved a lot (in circles).

Averages are worth calculating for the parameter space, the vector field of the cube. But not for individual countries, or particules moving within the cube. Be aware of that when presenting and interpreting the results. 
```{r fig.height=3,fig.width=3}
a1 <- df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$pam$`2`$clustering)) |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(EF, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = EF, yend = Gini, color = as.factor(Group)),
        arrow = arrow(length = unit(0.15, "cm")), size = 0.15) +
    geom_text(
        aes(x = EF *1.1, y = Gini * 1.1, label = country_code, colour = Group), size = 2) +
    scale_color_manual("Groups", values = c("orange", "purple")) +
    labs(x = "Ecological Footprint", y = "Gini") +
    theme_light(base_size = 6) +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2))

a1

c1 + annotation_custom(
        grob = ggplotGrob(a1),
        xmin = 10, xmax = 18, ymin = 40, ymax = 60
    ) 
```

```{r fig.width=2, fig.height=2}
a2 <- df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$pam$`2`$clustering)) |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(GNI, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = GNI, yend = Gini, 
            color = as.factor(Group)), 
        arrow = arrow(length = unit(0.15, "cm")), size = 0.15) +
    geom_text(aes(x = GNI * 1.1, y = Gini *1.1, 
                  label = country_code, colour = Group), size = 2) +
    scale_color_manual("Groups", values = c("orange", "purple")) +
    theme_light(base_size = 6) +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2))

c2 + annotation_custom(
        grob = ggplotGrob(a2),
        xmin = 1e+01, xmax = 1e+03, ymin = 40, ymax = 60
    )

a2
```
